---
title: "Sample POV Data Review"
author: "Jamie Robertson"
output:
  html_document: default
date: 'Jan. 2023'
subtitle: Capstone

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
options(show.signif.stars = FALSE)
library(knitr)
library(readr)
library(mosaic)
#library(effects)
library(graphics)
library(car)
library(MuMIn)

library(tidyverse,warn.conflicts = F)
library(tidymodels,warn.conflicts = F)
library(lubridate,warn.conflicts = F)
library(tidyr)
knitr::opts_chunk$set(echo = TRUE)
```


The data source: Poverello Center
Data dictionary for HMIS data: https://files.hudexchange.info/resources/documents/HMIS-Data-Dictionary.pdf


```{r show_col_types = FALSE}
#Read in the data

d.pov <- read_csv("C:\\Users\\jamie\\OneDrive\\Documents\\Capstone\\povdata.csv", show_col_types = FALSE)

```


Assumptions for LR:
1. independence
2. linearity of relationship
3. multicollinearity effects checked for
4. equal variance and normality of residuals
5. no influential points 


#Exploring the data
Looking at the first 6 rows of each database. 

```{r}

head(d.pov)

```

This analysis will review veteran status, primary and secondary race, ethnicity, length of stay, gender, if the client is insured, disabled, their date of birth and the date their homeless experience started, their date of birth, the first date of engagement with the system, the total number of months they have been homeless in the past three years, the number of times they have been on the streets or in the shelters in the past three years, their total monthly income, if they receive income from any source and if they receive a non-cash benefit from any source.


```{r}
d.pov <- d.pov %>% 
  mutate(totserv=efood_count+wfood_count+eother_count+wother_count)
  
```

Using Google Big Query, the count of services provided was summed up for each location and type. Two types of services are available for clients at two locations, Food or Basic Needs. Most individuals are receiving food. These counts were combined in R to be totserv which we will take a closer look at.

```{r}

hist(d.pov$totserv)


```

The distribution is heavily skewed. Another way to look at the distribution is using summary statistics and percentiles. 


```{r}

d.pov %>%
  pull(totserv) %>%
  summary()


```

```{r}
quantile(d.pov$totserv, probs = seq(.1, .9, by =.1))
```
```{r}
d <-
  d.pov %>%
  mutate(
    race = factor(race)
  )

d %>%
  pull(race) %>%
  summary()
```

Running a practice linear regression 

```{r}
# Remove rows where there is any missing data
#nhanes_subset_complete <- na.omit(nhanes_subset) 
library(MuMIn)
# Now refit the full model with only complete cases
fullModel <- lm(totserv ~ vet + race + eth +
                  gen + dvs + insured +   
                  los + disabled + any_income + 
                prior_situation, data=d.pov)
options(na.action="na.fail", width=120)
povOutput <- dredge(fullModel, 
       extra=c("R^2",adjRsq=function(x) summary(x)$adj.r.squared))
kable(head(povOutput,n=15),caption="Table 6: POV Table", digits=3)
```


