---
title: "Sample POV Data Review"
author: "Jamie Robertson"
output:
  html_document: default
date: 'Jan. 2023'
subtitle: Capstone

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
options(show.signif.stars = FALSE)
library(knitr)
library(readr)
library(mosaic)
#library(effects)
library(car)


library(tidyverse,warn.conflicts = F)
library(tidymodels,warn.conflicts = F)
library(lubridate,warn.conflicts = F)
library(tidyr)
knitr::opts_chunk$set(echo = TRUE)
```


The data source: Poverello Center
Data dictionary for HMIS data: https://files.hudexchange.info/resources/documents/HMIS-Data-Dictionary.pdf


```{r}
#Read in the data
d.soup <- read_csv("C:\\Users\\jamie\\OneDrive\\Documents\\Capstone\\SampleData-SoupKitchen.csv")
d.hmis <- read_csv("C:\\Users\\jamie\\OneDrive\\Documents\\Capstone\\SampleData-EmergencyShelter.csv")
```



Assumptions in the data, need to confirm 1&2 in Pov: 
1. d.soup is data collected The Pov
2. d.hmis is data collected by HMIS
3. Client ID is a primary key that can tie the data together. 

Assumptions for LR:
1. independence
2. linearity of relationship
3. multicollinearity effects checked for
4. equal variance and normality of residuals
5. no influential points 


#Exploring the data
Looking at the first 6 rows of each database. 

```{r, include=FALSE}

head(d.hmis)


```




```{r}
head(d.soup)
```

n_distinct(d.hmis$`Client ID`, na.rm = FALSE)
# 407

n_distinct(d.soup$'Client ID', na.rm = FALSE)
# 1488

```{r}

n_distinct(d.hmis$`Client ID`, na.rm = FALSE)
# 407

n_distinct(d.soup$'Client ID', na.rm = FALSE)
# 1488

```


